Fail Faster

### 2025-03-19 paper/paper.tex
flagging for better name

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Cynthia Richey}
\affiliation{%
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \state{Pennsylvania}
  \country{USA}
}

\author{Joseph W. Cutler}
\affiliation{
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \state{Pennsylvania}
  \country{USA}
}

@@

Property-based testing (PBT) is a software testing technique which uses
random test inputs to validate logical specifications.  Recent studies on PBT usage~\cite{inpractice} have shown that
practitioners often interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This means that PBT is often run with very short time
budgets---e.g., between 50 milliseconds and 30 seconds---and it is important
that bugs can be found in that time-frame.\iflater
\jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
how to fit this in yet}\fi

@@

However, careful measurements show that the performance of existing generator libraries falls significantly short of what
is possible. This is due to two major sources of inefficiency.
First, the same high-level design that gives generator libraries their
flexibility and expressiveness also introduces overhead: layered abstractions
and indirections hinder compiler optimizations, shifting work to
run time.
Second, the \rand that underlies the generator's operation---the source of the
random numbers that are used to make choices in the generator---can 
constitute an unexpectedly high proportion of a generator's running time.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi
Together, these observations raise a fundamental question: can generator libraries be expressive and high-level without compromising efficiency?
%In practice, PBT is used 
%as a lightweight, ad hoc approach to software correctness,
%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
%aimed at 
%detecting bugs within a short testing window---typically 50 milliseconds 
%to 30 seconds. \jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that 
%inputs be generated as fast as possible; in other words, generators must be 
%treated as performance-sensitive code.

% The locus for improvement in this area \jwc{``This area'' sounds like it's talking about the whole area of PBT... Maybe ``a high-leverage way to improve the performance of generators is to...''} is \textit{generator libraries}: if

  % in strict functionallanguages like OCaml and Scal
  To address this question, we outline two key principles for designing generator libraries that balance efficiency with flexibility.
 First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used to
optimize functional DSLs~\cn.
We apply this technique to generators, completely eliminating the overhead of
many common generator abstractions. For example, in unstaged generator
libraries, each call to monadic bind usually spends time constructing and
immediately executing a closure at run time. \jwc{Todo: joe --- cite some evidence for this}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the
compiler to inline and eliminate these closures.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \iflater\hg{More words here for
parallelism?}\fi Together, these insights provide a recipe for improving existing generator libraries.

%\jwc{I think we should use return/bind as the simplest example, since then you don't have to get into the weeds of allocation or closures. The functional abstractions of a monadic generator DSL block the compiler from generating fast code.}
%\jwc{Then present this case as ``more sophisticated use of staging'' or something?}
%where an unstaged generator DSL might represent a choice between
%a list of options as a dynamically-allocated array of closures, a staged 
%generator can expand to an if-else tree at compile 
%time---eliminating the allocation entirely.
% This approach is not tied to any
%particular PBT library, but rather provides a recipe for improving even
%state-of-the-art implementations \jwc{Sentence structure doesn't follow. Instead maybe ``not any particular, but a recipe for improving in general''}.

%First, we demonstrate the impact of
%RNG choice by optimizing an existing \rand. Our optimized RNG produces 
%exactly 
%the same values as its unoptimized counterpart, allowing us to directly 
%measure the impact of this choice on generator performance. \jwc{Permute? I think staging \emph{then} randomness.} \jwc{Also, i don't love this as the first place we explain the drop-in equivalence of the RNG. It's unmotivated why wen want this.}

@@

\section{Sources of Inefficiency in Monadic Generator Libraries}
\label{section:motiv}
\jwc{In this section we answer: ``What are monadic generator libraries, and why are they slow?''}
\bcp{That's actually not a bad section title!}

@@

In the years since the original QuickCheck paper \cn{}, PBT has had remarkable success
in languages outside of the small ICFP world from which it came.
For this reason, readers who are users and developers of PBT libraries in \emph{other} languages ---
including non-strict or procedural ones --- might be curious about how the
techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper
might be imported to their favorite language.
Below we discuss how \name{} could be applied in other languages.

@@

  \item \textbf{Haskell} birthed PBT, and so it is natural to ask why we have not yet applied \name{} to it. The basic answer is that
  the Haskell compiler (GHC) is designed specially to eliminate the run-time overhead of monadic
  abstractions! Indeed, because Haskell is pure, GHC can aggressively inline and beta-reduce
  nearly anything as part of its normal optimiation steps.
  This means that in many cases, the impact of a staged PBT library (and indeed, all staged monadic DSLs) negligible.
  For complex enough programs however, the heuristics GHC uses can degrade, leading to performance overheads for monadic code.
  For this reason, Haskell does have a multi-stage programming system called Template Haskell \cn{}, which is sometimes 
  to build eDSLs in contexts where one does not want to rely on the optimizer's heuristics.
  In short: it is possible to port \name{} to Haskell QuickCheck, but the benefits would be much less pronounced.

>>>>

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Cynthia Richey}
\authornote{Both authors contributed equally to this work.}
\affiliation{%
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \state{Pennsylvania}
  \country{USA}
}

\author{Joseph W. Cutler}
\authornotemark[1]
\affiliation{
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \state{Pennsylvania}
  \country{USA}
}

@@

Property-based testing (PBT) is a software testing technique which uses
random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} have shown that
practitioners often interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This means that PBT is often run with very short time
budgets---e.g., between 50 milliseconds and 30 seconds---and it is important
that bugs can be found in that time-frame.\iflater
\jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
how to fit this in yet} \jwc{Also think the phrasing should say something like ``the goal is to fail as fast as possible''}\fi

@@

However, careful measurements show that the performance of existing generator libraries falls significantly short of what
is possible. This is due to two major sources of inefficiency.
First, the same high-level design that gives generator libraries their
flexibility and expressiveness also introduces overhead: layered abstractions
and indirections hinder compiler optimizations, shifting work to
run time.
Second, the \rand that underlies the generator's operation---the source of the
random numbers that are used to make choices in the generator---can 
constitute an unexpectedly high proportion of a generator's running time.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi
Together, these observations raise a fundamental question: can generator libraries be expressive and high-level without compromising efficiency?

  To address this question, we outline two key principles for designing generator libraries that balance efficiency with flexibility.
 First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used to
optimize functional DSLs~\cn.
We apply this technique to generators, completely eliminating the overhead of
many common generator abstractions. For example, in unstaged generator
libraries, each call to monadic bind usually spends time constructing and
immediately executing a closure at run time. \jwc{Todo: joe --- cite some evidence for this}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the
compiler to inline and eliminate these closures.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \iflater\hg{More words here for
parallelism?}\fi Together, these insights provide a recipe for improving existing generator libraries.

@@

\section{Sources of Inefficiency in Monadic Generator Libraries}
\label{section:motiv}
\jwc{In this section we answer: ``What are monadic generator libraries, and why are they slow?''}
\bcp{That's actually not a bad section title!}
\todo{Intro of this section}

@@

In the years since the original QuickCheck paper \cn{}, PBT has had remarkable success
in languages outside of the Haskell world from which it came.
For this reason, readers who are users and developers of PBT libraries in \emph{other} languages ---
including non-strict or procedural ones --- might be curious about how the
techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper
might be imported to their favorite language.
Below we discuss how \name{} could be applied in other languages.

@@

  \item \textbf{Haskell} is home to the original PBT implementation \cn{}, and
  so it is natural to ask why we have not yet applied \name{} to it. The basic
  answer is that the Haskell compiler (GHC) is designed specially to eliminate
  the run-time overhead of monadic abstractions! Indeed, because Haskell is
  pure, GHC can aggressively inline and beta-reduce nearly anything as part of
  its normal optimiation steps.  This means that in many cases, the impact of a
  staged PBT library (and indeed, all staged monadic DSLs) negligible.  For
  complex enough programs however, the heuristics GHC uses can degrade, leading
  to performance overheads for monadic code.  For this reason, Haskell does have
  a multi-stage programming system called Template Haskell \cn{}, which is
  sometimes to build eDSLs in contexts where one does not want to rely on the
  optimizer's heuristics.  In short: it is possible to port \name{} to Haskell
  QuickCheck, but the benefits would be much less pronounced.
